---
title: "Assignment3-47506067"
author: "Imtiaz Ahmed"
date: "2023-10-28"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, warning=FALSE, include=FALSE}
library(ISwR)
library(tidyverse)
library(distr)
library(GGally)
library(car)
library(ggplot2)
library(dplyr)
library(patchwork)
library(gamlss)
library(ggmosaic)
library(janitor)
library(VGAM)
library(geepack)
```


```{r echo=FALSE, warning=FALSE}
doc_data <- read.csv('doc_visits.csv', header = TRUE)

```

Q1(a)

```{r echo=FALSE, warning=FALSE}
ggplot(doc_data) +
geom_bar(aes(x = doc_visits))

```

```{r echo=FALSE, warning=FALSE}
ggplot(doc_data) +
geom_bar(aes(x=doc_visits)) +
facet_grid(cols = vars(age50))
```
The count plots convincingly show that there are a lot of zero values for doctor visits. Especially for age<50 group. A zero-inflated models should be considered for this case.

Q1(b)

```{r echo=FALSE, warning=FALSE}
par(mfrow=c(1,2))
p1<-ggplot(data.frame(doc_data), aes(x=schooling)) +
  geom_histogram(binwidth=3, fill="blue") +
  labs(title="Histogram of schooling", x="schooling") +
  theme_minimal()
p2<-ggplot(data.frame(doc_data), aes(sample = schooling)) +
  geom_qq() +
  geom_qq_line(color = "red", lwd = 2) +
  labs(title = "Q-Q Plot of schooling") +
  theme_minimal()
p1+ p2
```

Schooling looks to be right skewed. The q-q plot shows similar issue with the curve.
```{r echo=FALSE, warning=FALSE}
GGally::ggpairs(doc_data)
```

There appears to be negative correlation between health and doc_visits. as intuitively understood, high health satisfaction score results in lower doc_visits.age<50 overall has less doc_visits compared to age>50.
```{r echo=FALSE, warning=FALSE}
model <- gamlss(doc_visits ~ health + age50 + schooling, family = "ZINBI", data = doc_data)
```

```{r echo=FALSE, warning=FALSE}
best_model <- stepGAIC(model, method="backward")
```

```{r echo=FALSE, warning=FALSE}
summary(best_model)
```

```{r echo=FALSE, warning=FALSE}
plot(best_model)
```

The quantile residuals appear to be normally distributed so the model is good.
Fitted Model Equations:
$$ Y_i \sim ZINB(\mu_i,\sigma,\nu_i),  i = 1,...,300 $$
$$ log(\mu_i) = 2.56961 -0.26623x_i1 + 0.42067x_i2 + log(u_i) $$
Where, $x_i1$ is health, $x_i2$ is 1 for age50=True, 0 for age50=False.
$$ \sigma = e^(0.2165) = 1.24 $$
$$ log(\pi_i\div(1-\pi_i)) = -36.04 $$

Parameter Interpretation:
The effect on mean number of doctor visits, for age more than 50 is exp(0.42067) = 1.52. So having age more than 50 increase the number of doctor visits by 52%.
The effect on mean number of doctor visits, for 1 unit increase in health score is exp(-0.26623) = 0.766. So with each unit increase in health score decreases the number of doctor visits by 23%.

Q2(a)

```{r echo=FALSE, warning=FALSE}
crash_data <- read.csv('crash.csv', header = TRUE)
```

```{r echo=FALSE, warning=FALSE}
breaks <- c(135, 520, 900, 1255, 1575, 1860, Inf)
labels <- c("1", "2", "3", "4", "5", "6")
crash_data$ais <- cut(crash_data$head_ic, breaks = breaks, labels = labels, include.lowest = TRUE)
table(crash_data$ais, useNA = "ifany")
```
It seems there are 12 missing values in the ais column.Category 5 and 6 have less than 20 observations. We will drop the observations with missing values and combine category 5 and 6 into 5.
```{r echo=FALSE, warning=FALSE}
crash_data <- na.omit(crash_data)
crash_data$ais <- as.factor(as.character(crash_data$ais))
levels(crash_data$ais)[levels(crash_data$ais) == "6"] <- "5"
table(crash_data$ais, useNA = "ifany")
```

Q2(b)

```{r echo=FALSE, warning=FALSE}
ggplot(crash_data) +
geom_mosaic(aes(x = product(ais, dp)), show.legend = FALSE) +
xlab("dp") +
ylab("ais")
```

```{r echo=FALSE, warning=FALSE}
ordinal_model <- vglm(ordered(ais) ~ factor(dp) + weight,
family = cumulative(parallel = TRUE), data = crash_data
)
summary(ordinal_model)
```

Model equation:
Estimated model equation for j = 1 is,
$$ ln(\gamma_1i/(1 - \gamma_1i)) = -0.7162 + 0.8800x_i1 - 0.00045x_i2 $$
Where, j denotes the ordinal level of ais, $x_i1$ is 1 for dp=Passen, 0 for dp=Driver, $x_i2$ is weight.
Parameter Interpretation:
We have exp(0.88) = 2.41. So, for any ais group j = 1, 2, 3, 4 the
odds of being in group less than j for passenger is 2.41 times the odds of being in group less than j for driver,
for cars of same weight. This is an increase in the odds of 241%.
We have exp(-0.00045) = 0.99. So, for every unit increase in weight, there
is a 1% reduction in the odds of being in ais group less than j, for j = 1, 2, 3, 4.

Q2(c)

```{r echo=FALSE, warning=FALSE}
nominal_model <- vglm(ais ~ factor(dp) + weight,
family = multinomial(refLevel = "1"), data = crash_data
)
summary(nominal_model)
```
Model Equation for ais = 2:
$$ ln(\pi_2i/\pi_1i) = 1.4317 - 1.5372x_i1 - 0.0001x_i2  $$
Where, $x_i1$ is 1 for Passenger, 0 for driver, $x_i2$ is weight.
Parameter Interpretation:
the effect on the odds of linear fracture, relative to Headache or dizziness, of a one unit increase in weight,
is exp(-0.0001) = 0.99, i.e. a 1% decrease in the odds of linear fracture with every unit increase in weight.
the effect on the odds of linear fracture, relative to headache or dizziness, of being in passenger seat compared with being in driver seat, is exp(-1.5372)= 0.21 i.e. a 79%% decrease in the odds of linear fracture.

Q3(a)

```{r echo=FALSE, warning=FALSE}
load("data.child.Rdata")
```


```{r echo=FALSE, warning=FALSE}
child_gee_exch <- geeglm(Ht ~ Agedays + factor(Sex),
id = Id,
corstr = "exchangeable", data = data.child, family = gaussian()
)
child_gee_exch2 <- geeglm(Ht ~ factor(Sex) + Agedays + I(Agedays^2),
id = Id,
corstr = "exchangeable", data = data.child, family = gaussian()
)
child_gee_exch3 <- geeglm(Ht ~ factor(Sex) + Agedays + I(Agedays^2) + I(Agedays^3),
id = Id,
corstr = "exchangeable", data = data.child, family = gaussian()
)
QIC(child_gee_exch, child_gee_exch2, child_gee_exch3 )
```
```{r echo=FALSE, warning=FALSE}
anova(child_gee_exch3)
```
All the covariates come out highly significant in the model child_gee_exch3.

Q3(b)

The response height is assumed to be normally distributed. Looking at the histogram also confirms that this assumption is fair.

```{r echo=FALSE, warning=FALSE}
ggplot(data.frame(data.child), aes(x=Ht)) +
  geom_histogram(binwidth=10, fill="blue") +
  labs(title="Histogram of height", x="Height") +
  theme_minimal()
```

Q3(c)

```{r echo=FALSE, warning=FALSE}
summary(child_gee_exch3)
```
Fitted Model Equation:
$$ log(\mu_ij) = 61.7 + 1.54s_j + 0.101x_i1 - 0.000134x_i2 + 0.0000000836x_i3$$
Where, $s_j$ is 1 for female, 0 for male, $x_i1, x_i2, x_i3$ are the base, quadratic and cubic terms for Agedays respectively.

Q3(d)

The working correlation matrix $$ R_i = \begin{pmatrix}   1 & 0.66 & 0.66 \\ . & 1 & . \\ 0.66 & 0.66 & 1 \\ \end{pmatrix} $$
And the correlation matrix is given by $$ R = \begin{pmatrix}   R_i & 0 & 0 \\ . & R_i & . \\ 0 & 0 & R_i \\ \end{pmatrix} $$

Q3(e)

The effect of age is cubic on height and highly significant. (p less than 2e-16)
beta = 1.54 shows that females have an increase in height compared to males child.

Q3(f)

Since the data is longitudinal observations, ar1 correlation structure seems to be a reasonable case.We create a model with ar1 correlation structure and use QIC to compare.
```{r echo=FALSE, warning=FALSE}
child_gee_ar1 <- geeglm(Ht ~ factor(Sex) + Agedays + I(Agedays^2) + I(Agedays^3),
id = Id,
corstr = "ar1", data = data.child, family = gaussian()
)
QIC(child_gee_exch3, child_gee_ar1)
```
Previous model with exchange structure appears to be the better model based on QIC values.

Q3(g)

Linear Mixed Model:
$$ Y_ij = \beta_0 + \beta_1 age + \beta_2 age^2 + \beta_3 age^3 + \gamma_1 sex1 + b_i + \epsilon_ij $$ 
Where, $$ \beta_0  = intercept, $$
$$ \beta_1, \beta_2, \beta_3  are growth over time (slope) $$
$$ b_i  \sim N(0, \nu^2)  independently $$
$$ \epsilon_ij \sim N(0, t^2)  independently $$

Q3(j)

```{r echo=FALSE, warning=FALSE}
plot <- ggplot(data.child, aes(x = Agedays, y = Ht)) +
geom_point() +
geom_smooth(method = "lm",
formula = y ~ splines::bs(x, degree = 5),
se = FALSE)
plot
```

Q3(k)

Looking at the scatterplot, the linear growth model seems to be appropriate because it seems like for each child, the 
growth is approximately linear.

Q3(l) 

There is a bit of curve in the spline, this suggests higher order form being present in the model. so Agedays 
should be tried in quadratic and cubic form in the model.
