---
title: "Assignment2"
author: "Imtiaz Ahmed - 47506067"
date: "2023-10-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r echo=FALSE, warning=FALSE, include=FALSE}
library(ISwR)
library(tidyverse)
library(distr)
library(GGally)
library(car)
library(ggplot2)
library(dplyr)
library(patchwork)
```
Q1(a)
```{r echo=FALSE, warning=FALSE}
pbc_data <- read.csv('pbc.csv', header = TRUE)
ggplot(data.frame(pbc_data), aes(x=serBilir)) +
  geom_histogram(binwidth=10, fill="blue") +
  labs(title="Histogram of serBilir", x="serBilir") +
  theme_minimal()

```
The histogram plot shows right skewness. Such data is modelled better with gamma or inverse gaussion distribution since those distributions are right skewed and fits the data well.

Q1(b)
```{r echo=FALSE, warning=FALSE}
par(mfrow=c(1,2))
p1<-ggplot(data.frame(pbc_data), aes(x=alkaline)) +
  geom_histogram(binwidth=1000, fill="blue") +
  labs(title="Histogram of alkaline", x="alkaline") +
  theme_minimal()
p2<-ggplot(data.frame(pbc_data), aes(sample = alkaline)) +
  geom_qq() +
  geom_qq_line(color = "red", lwd = 2) +
  labs(title = "Q-Q Plot of alkaline") +
  theme_minimal()
p1+ p2
```
The histogram plot on Alkaline shows right skewness. Log or sqrt transformations are used for dealing with right skewness.

 The qq plot shows extreme values on the right side. This can be mitigated by log transform because log transform penalizes extreme values. sqrt transform will also work for the same reasons.
```{r echo=FALSE, warning=FALSE}
par(mfrow=c(1,2))
p3<-ggplot(data.frame(pbc_data), aes(x=prothrombin)) +
  geom_histogram(binwidth=2, fill="blue") +
  labs(title="Histogram of prothrombin", x="prothrombin") +
  theme_minimal()
p4<-ggplot(data.frame(pbc_data), aes(sample = prothrombin)) +
  geom_qq() +
  geom_qq_line(color = "red", lwd = 2) +
  labs(title = "Q-Q Plot of prothrombin") +
  theme_minimal()
p3+ p4
```

There's right skewness in prothrombin. log or sqrt transform can help normalize the data.

qqplot of prothrombin shows curve. log or sqrt transform can resolve the issue.
```{r echo=TRUE, warning=FALSE, fig.show='hide'}
pbc_data$logalkaline<- log(pbc_data$alkaline)
pbc_data$logprothrombin <- log(pbc_data$prothrombin)
pbc_data$logserBilir <- log(pbc_data$serBilir)
```
```{r echo=FALSE, warning=FALSE}
par(mfrow=c(1,2))
p1<-ggplot(data.frame(pbc_data), aes(x=logalkaline)) +
  geom_histogram(binwidth=0.1, fill="blue") +
  labs(title="Histogram of logalkaline", x="logalkaline") +
  theme_minimal()
p2<-ggplot(data.frame(pbc_data), aes(x=logprothrombin)) +
  geom_histogram(binwidth=0.1, fill="blue") +
  labs(title="Histogram of logprothrombin", x="logprothrombin") +
  theme_minimal()
p1+p2
```



Q1(c)
```{r echo=FALSE, warning=FALSE}
GGally::ggpairs(pbc_data[, c("age", "albumin", "logalkaline", "logprothrombin","logserBilir")])
```

logalkaline, logprothrombin have positive correlation with response, albumin has negative correlation with response. All three of these continuous variables can be predictors in a statistical model.However, albumin has negative correlation with logalkaline and logprothrombin. So albumin should not be in the model if the other two are present.

Q1(d)
```{r echo=FALSE, warning=FALSE}
p1 <- table(pbc_data$sex)
p2 <- table(pbc_data$hepatomegaly)
p3 <- table(pbc_data$histologic)
list(sex=p1, hepatomegaly=p2, histologic=p3)
```
It appears there's a lot less data for males compared to females. hepatomegaly has fairly similar number of data for each level. There's very few patients with lowest stage of the disease.

Q1(e)
```{r echo=FALSE, warning=FALSE}
par(mfrow=c(1,3))
boxplot(logserBilir ~ sex, data=pbc_data, main="Boxplot of logserBilir by sex")
boxplot(logserBilir ~ hepatomegaly, data=pbc_data, main="Boxplot of logserBilir by hepatomegaly")
boxplot(logserBilir ~ histologic, data=pbc_data, main="Boxplot of logserBilir by histologic")
par(mfrow=c(1,1))
```
serBilir does not seem to be affected much by sex. males have a slightly higher mean but that can be attributed to lower data points compared to females. So sex may not be a predictor for the response.
hepatomegaly seems to affect serBilir levels looking at the boxplot. patients with enlarged liver have higher levels of serum bilirubin. This is a potential predictor of the response.
Looking at the boxplot for histologic, we can see a trend. The more severe stage of disease, the higher levels of serum bilirubin. hostologic can be a predictor of serBilir.


Q1(f)
```{r echo=FALSE, warning=FALSE}
pbc_data$histologic <- factor(pbc_data$histologic)
pbc_data$sex <- factor(pbc_data$sex)
pbc_data$hepatomegaly <- factor(pbc_data$hepatomegaly)
results <- data.frame(Model = character(), P_Value = numeric(), Scaled_Deviance = numeric(), AIC = numeric())

  model <- glm(formula = serBilir ~ histologic, family=Gamma(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[4, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "histologic", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ hepatomegaly, family=Gamma(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "hepatomegaly", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ sex, family=Gamma(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "sex", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ age, family=Gamma(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "age", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ albumin, family=Gamma(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "albumin", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ logalkaline, family=Gamma(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "logalkaline", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ logprothrombin, family=Gamma(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "logprothrombin", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
print(results)

```
Q1(g)
```{r echo=FALSE, warning=FALSE}
results <- data.frame(Model = character(), AIC = numeric())
model <- glm(formula = serBilir ~ histologic, family=Gamma(link="log"), data=pbc_data)
aic_val <- AIC(model)
results <- rbind(results, data.frame(Model = "histologic", AIC = aic_val))
model1 <- glm(formula = serBilir ~ histologic + logprothrombin, family=Gamma(link="log"), data=pbc_data)
aic_val <- AIC(model1)
results <- rbind(results, data.frame(Model = "histologic+logprothrombin", AIC = aic_val))
model2 <- glm(formula = serBilir ~ histologic + logprothrombin + logalkaline, family=Gamma(link="log"), data=pbc_data)
aic_val <- AIC(model2)
results <- rbind(results, data.frame(Model = "histologic+logprothrombin+logalkaline", AIC = aic_val))
model3 <- glm(formula = serBilir ~ histologic + logprothrombin + logalkaline + hepatomegaly, family=Gamma(link="log"), data=pbc_data)
aic_val <- AIC(model3)
results <- rbind(results, data.frame(Model = "histologic+logprothrombin+logalkaline+hepatomegaly", AIC = aic_val))
model4 <- glm(formula = serBilir ~ histologic + logprothrombin + logalkaline + hepatomegaly + albumin, family=Gamma(link="log"), data=pbc_data)
aic_val <- AIC(model4)
results <- rbind(results, data.frame(Model = "histologic+logprothrombin+logalkaline+hepatomegaly+albumin", AIC = aic_val))
print(results)
```
We start with only histologic as the covariate which gives us an AIC of 1325.9. We add logprothrombin to the model and obtain AIC 1276.5. Then we add logalkaline to the model and obtain AIC score of 1241.2. Adding hepatomegaly to the model gives an AIC of 1216.2. Lastly, we add albumin to the model and obtain AIC score of 1194.7. So the last model is our final model.

Q1(h)
```{r echo=FALSE, warning=FALSE}
summary(model4)
```
serBilir ~ Gamma($\alpha$, $\beta$)
The model equation is:
$$ log(serBilir) = \beta_0 + \beta_1*histologic2 + \beta_2*histologic3 + \beta_3*histologic4 $$
$$ + \beta_4*logprothrombin + \beta_5*logalkaline + \beta_6*hepatomegalyYes + \beta_7*albumin $$
Where,

histologic2, histologic3 and histologic4 are dummy variables.If histologic is at level 2, histologic2 will be 1 and others 0. if histologic is at level 3, histologic3 will be 1 and others 0. if histologic is at level 4, hostologic4 will be one and others 0. If histologic is at level 1, histologic2, histologic3 and histologic4 will be 0.

Similarly, hepatomegalyYes is a dummy variable. It is 1 for enlarged liver, 0 for not.

serBilir is the response variable and remaining covariates are continuous variables of the same name.

$$ \beta_0 = -8.17804 , \beta_1 = 0.35606, \beta_2 = 0.49777, \beta_3 = 0.44692, \beta_4 = 3.16285, \beta_5 = 0.40733, \beta_6 = 0.48293, \beta_7 = -0.57146 $$
The final model equation is thus,
$$ log(serBilir) = -8.17804 + 0.35606*histologic2 + 0.49777*histologic3 + 0.44692*histologic4 + 3.16285*logprothrombin $$
$$ + 0.40733*logalkaline + 0.48293*hepatomegalyYes - 0.57146*albumin $$

Q1(i)
Since, histologic is not significant in the final model, we will not do parameter interpretations for those dummy variables.

a 10% increase in prothrombin has the effect of $1.1^(3.16285)=1.3518$ increase in serum bilirubin or a 35% increase

a 10% increase in alkaline phosphatase has the effect of $1.1^(0.40377)=1.0392$ increase in serum bilirubin or a 4% increase.

The multiplicative effect on expected serum bilirubin of having enlarged liver compared with
normal, is $e^(0.48293)=1.6208$ time increase or 62% increase

The multiplicative effect on expected serum bilirubin of an increase in 1 gm/dl in albumin is $e^(-0.57146)=0.5647$ times decrease or 44% decrease

Q1(j)
Patients who have enlarged liver, whos disease stage has progressed to advanced levels, have higher amount of prothrombin and alkaline exibit higher levels of serum bilirubin.
Q1(k)
Check for model fit:
we check scaled deviance against n-p which in this case, 304.
```{r echo=TRUE, warning=FALSE}
par(mfrow=c(1,1))
scaled_deviance = model4$deviance/summary(model4)$dispersion
scaled_deviance
```
So no problem with the scaled deviance. Now we check the quantile residuals,
```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align='left'}
library(DHARMa)
library(ggplot2)
options(ggrepel.max.overlaps = Inf)
res_model4 <- simulateResiduals(fittedModel = model4)
q_model4 <- residuals(res_model4, quantileFunction = qnorm)
p1 <- ggplot(data = tibble(q = q_model4), aes(x = q_model4)) +
geom_histogram(aes(y = after_stat(density)), bins = 6) +
geom_density()
p1
```
The distribution of the residuals look right skewed. Let's do more checks,
```{r echo=FALSE, warning=FALSE, fig.align='left'}
par(mfrow=c(1,1))
plot(res_model4)
```
Q-Q plot residuals show significant result on KS test. residuals vs predicted plot shows significant quantile deviation. The model is not a good fit.

Mis-specification of the systematic part of the model: We check partial residual plots for the covariates,

```{r echo=FALSE, warning=FALSE, fig.align='left'}
par(mfrow=c(1,1))
crPlots(model4)
```

logprothrombin and logalkaline seems to have slight non-linearity.Rest of the plots look okay.
we can say systematic part of the model is specified correctly.

Unusual observations: 
```{r echo=FALSE, warning=FALSE, fig.align='left'}
diag4 <- broom::augment(model4)
p1<-ggplot(diag4) +
geom_point(aes(y = .resid, x = seq_along(.resid))) +
xlab("Index") + ylab("Deviance Residuals")
```
The deviance residual plot shows a random scatter which is good. The cut off value for leverage in this case is 0.05
```{r echo=FALSE, warning=FALSE, fig.align='left'}
par(mfrow=c(1,2))
p2<-ggplot(diag4, aes(y = .hat, x = seq_along(.hat))) +
geom_point() +
xlab("Index") +
ylab("Leverage") +
geom_hline(yintercept = 0.05, linetype = 2) +
ggrepel::geom_text_repel(aes(label = seq_along(.hat)))

p3<-ggplot(diag4, aes(y = .cooksd, x = seq_along(.hat))) +
geom_point() +
xlab("Index") +
ylab("Cook’s Distance") +
ggrepel::geom_text_repel(aes(label = seq_along(.hat)))
p1+p2+p3
```
There are a lot of points with high leverage. Lastly, we check cook's distance to determine if any of these high leverage points have out-of-proportion effect on the coefficients.
No points have cook's distance greater than 1. So we conclude no unusual observations.
Q1(l)
```{r echo=FALSE, warning=FALSE, results='hide'}
results <- data.frame(Model = character(), P_Value = numeric(), Scaled_Deviance = numeric(), AIC = numeric())

  model <- glm(formula = serBilir ~ histologic, family=inverse.gaussian(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[4, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "histologic", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ hepatomegaly, family=inverse.gaussian(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "hepatomegaly", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ sex, family=inverse.gaussian(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "sex", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ age, family=inverse.gaussian(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "age", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ albumin, family=inverse.gaussian(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "albumin", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ logalkaline, family=inverse.gaussian(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "logalkaline", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm(formula = serBilir ~ logprothrombin, family=inverse.gaussian(link="log"), data=pbc_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "logprothrombin", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
print(results)
```

```{r echo=FALSE, warning=FALSE, results='hide'}
results <- data.frame(Model = character(), AIC = numeric())
model <- glm(formula = serBilir ~ histologic, family=inverse.gaussian(link="log"), data=pbc_data)
aic_val <- AIC(model)
results <- rbind(results, data.frame(Model = "histologic", AIC = aic_val))
model1 <- glm(formula = serBilir ~ histologic + logprothrombin, family=inverse.gaussian(link="log"), data=pbc_data)
aic_val <- AIC(model1)
results <- rbind(results, data.frame(Model = "histologic+logprothrombin", AIC = aic_val))
model2 <- glm(formula = serBilir ~ histologic + logprothrombin + logalkaline, family=inverse.gaussian(link="log"), data=pbc_data)
aic_val <- AIC(model2)
results <- rbind(results, data.frame(Model = "histologic+logprothrombin+logalkaline", AIC = aic_val))
model3 <- glm(formula = serBilir ~ histologic + logprothrombin + logalkaline + hepatomegaly, family=inverse.gaussian(link="log"), data=pbc_data)
aic_val <- AIC(model3)
results <- rbind(results, data.frame(Model = "histologic+logprothrombin+logalkaline+hepatomegaly", AIC = aic_val))
model4 <- glm(formula = serBilir ~ histologic + logprothrombin + logalkaline + hepatomegaly + albumin, family=inverse.gaussian(link="log"), data=pbc_data)
aic_val <- AIC(model4)
results <- rbind(results, data.frame(Model = "histologic+logprothrombin+logalkaline+hepatomegaly+albumin", AIC = aic_val))
print(results)
```
Diagnostic checks:
```{r echo=FALSE, warning=FALSE}
par(mfrow=c(1,1))
summary(model4)
scaled_deviance = model4$deviance/summary(model4)$dispersion
scaled_deviance
```
scaled deviance 273.1698 is okay against 304 degrees of freedom.
```{r echo=FALSE, warning=FALSE, fig.align='left'}

res_model4 <- simulateResiduals(fittedModel = model4)
q_model4 <- residuals(res_model4, quantileFunction = qnorm)
p1 <- ggplot(data = tibble(q = q_model4), aes(x = q_model4)) +
geom_histogram(aes(y = after_stat(density)), bins = 6) +
geom_density()
p1
```
quantile residuals look normally distributed

```{r echo=FALSE, warning=FALSE, fig.align='left'}
plot(res_model4)
```
The QQ plot does not show any significant results which is good for model fit.Residual vs predicted plot also does not detect any significant problem so the model is a good fit.
```{r echo=FALSE, warning=FALSE, fig.align='left'}
crPlots(model4)
```
All the variables look fairly linear. We can say systematic part of the model is specified correctly.
```{r echo=FALSE, warning=FALSE, fig.align='left'}
diag4 <- broom::augment(model4)
p1<-ggplot(diag4) +
geom_point(aes(y = .resid, x = seq_along(.resid))) +
xlab("Index") + ylab("Deviance Residuals")
```
Deviance residuals are a random scatter. no obvious pattern which is good.
```{r echo=FALSE, warning=FALSE, fig.align='left'}
par(mfrow=c(1,2))
p2<-ggplot(diag4, aes(y = .hat, x = seq_along(.hat))) +
geom_point() +
xlab("Index") +
ylab("Leverage") +
geom_hline(yintercept = 0.05, linetype = 2) +
ggrepel::geom_text_repel(aes(label = seq_along(.hat)))

p3<-ggplot(diag4, aes(y = .cooksd, x = seq_along(.hat))) +
geom_point() +
xlab("Index") +
ylab("Cook’s Distance") +
ggrepel::geom_text_repel(aes(label = seq_along(.hat)))
p1+p2+p3
```
We see quite a few high leverage points.
No cook's distance >1 so we conclude no unusual observation. 
This model was created using the same procedure as the gamma model. We did single variable regressions and discarded age, sex due to p values. Then we do forward model selection and end up with the same model specification as before glm(formula = serBilir ~ histologic + logprothrombin + logalkaline + hepatomegaly + albumin, family=inverse.gaussian(link="log"), data=pbc_data)
But with a lower AIC value of 1115.499. This model fares much better than the previous model when we do diagnostic check.


Q2(a)
```{r echo=FALSE, warning=FALSE}
hd_data <- read.csv('Heart_Disease.csv', header = TRUE)
p1<-ggplot(hd_data, aes(x = ER_visits)) +
  geom_histogram(aes(y = ..count..),
                 breaks = seq(0, max(hd_data$ER_visits)+5, by = 1),
                 fill = "#66c2a5", 
                 color = "white",) +
  labs(title = "Histogram of ER_visits",
       x = "ER_visits",
       y = "Frequency") +
  theme_minimal()

lambda <- mean(hd_data$ER_visits)
simulated_data <- rpois(length(hd_data$ER_visits), lambda)

p2<-ggplot(data.frame(simulated_data), aes(x = simulated_data)) +
  geom_histogram(binwidth = 1, fill = "#fc8d62", color = "white") +
  labs(title = "Simulated Poisson Dist lamda 3.42",
       x = "ER Visits",
       y = "Frequency") +
  theme_minimal()
p1 + p2
```

THe histogram of ER_visits and a simulated histogram based on the mean of ER_visits that follow poisson distribution show signs of overdispersion in the data. We see lower than expected frequency around the mean and more frequency in other places.

Q2(b)
```{r echo=FALSE, warning=FALSE, fig.align='left'}
par(mfrow=c(1,2))
p1<-ggplot(data.frame(hd_data), aes(x=ER_visits)) +
  geom_histogram(binwidth=1, fill="blue") +
  labs(title="Histogram of ER_visits", x="ER_visits") +
  theme_minimal()
p2<-ggplot(data.frame(hd_data), aes(sample = ER_visits)) +
  geom_qq() +
  geom_qq_line(color = "red", lwd = 2) +
  labs(title = "Q-Q Plot of ER_visits") +
  theme_minimal()
p1+ p2
```
While we see skewness in the ER_visits data, it is the response variable and will use a link function (likely log) anyway so we don't need to transform ER_visits
```{r echo=FALSE, warning=FALSE, fig.align='left'}
par(mfrow=c(1,2))
p1<-ggplot(data.frame(hd_data), aes(x=Age)) +
  geom_histogram(binwidth=1, fill="blue") +
  labs(title="Histogram of Age", x="Age") +
  theme_minimal()
p2<-ggplot(data.frame(hd_data), aes(sample = Age)) +
  geom_qq() +
  geom_qq_line(color = "red", lwd = 2) +
  labs(title = "Q-Q Plot of Age") +
  theme_minimal()
p1+ p2
```
Age is left skewed in the data, we will not transform this data because in the context of count regression, in does not make sense in terms of interpretability to normalize Age. Age is also a count variable so transformations will make it hard to understand resulting models.
```{r echo=FALSE, warning=FALSE, fig.align='left'}
par(mfrow=c(1,2))
p1<-ggplot(data.frame(hd_data), aes(x=Total_Cost)) +
  geom_histogram(binwidth=10000, fill="blue") +
  labs(title="Histogram of Total_cost", x="Total_cost") +
  theme_minimal()
p2<-ggplot(data.frame(hd_data), aes(sample = Total_Cost)) +
  geom_qq() +
  geom_qq_line(color = "red", lwd = 2) +
  labs(title = "Q-Q Plot of Total_cost") +
  theme_minimal()
p1+ p2
```
Total_Cost variable is extremely right skewed. It is worthwhile to use log or sqrt transformation on this. Total_cost is not a count variable so we have no issue here.
```{r echo=FALSE, warning=FALSE, fig.align='left'}
par(mfrow=c(1,2))
p1<-ggplot(data.frame(hd_data), aes(x=Comorbidities)) +
  geom_histogram(binwidth=1, fill="blue") +
  labs(title="Histogram of Comorbidities", x="Comorbidities") +
  theme_minimal()
p2<-ggplot(data.frame(hd_data), aes(sample = Comorbidities)) +
  geom_qq() +
  geom_qq_line(color = "red", lwd = 2) +
  labs(title = "Q-Q Plot of Comorbidities") +
  theme_minimal()
p1+ p2
```
Coborbidities is right skewed and is not a count variable. It can be transformed using log or sqrt to normalize.
```{r echo=TRUE, warning=FALSE}
hd_data$logtotal_cost <- log(hd_data$Total_Cost + 1)
hd_data$logcomorbidities <- log(hd_data$Comorbidities + 1)
```
Q2(c)
```{r echo=TRUE, warning=FALSE, fig.align='left'}
GGally::ggpairs(hd_data[, c("Age", "logtotal_cost", "Interventions", "Drug","Complications", "logcomorbidities", "Duration", "ER_visits")])
```

Drug is likely to be predictor for ER_visits. Interventions and logtotal_cost can also be predictors. Duration and logcomorbidities are correlated but none of them seem to be good predictors for ER_visits.
```{r echo=TRUE, warning=FALSE, fig.align='left'}
boxplot(ER_visits ~ Gender, data=hd_data, main="Boxplot of ER_visits by Gender")
```
Categorical variable Gender does not appear to be a predictor for ER_visits.So Drug, Interventions and logtotal_cost are only three predictors of ER_visits.
Q2(d)
```{r echo=TRUE, warning=FALSE}
model <- glm(formula = ER_visits ~ Drug , family=poisson(link = 'log'), data=hd_data)
summary(model)
```
Looking at the model summary, we see residual deviance quite a bit larger than 786 degrees of freedom. This indicates overdispersion as previously seen in the histograms. negative binomial GLM will be a better model in this case because negative binomial does not have the restriction of $\lambda=E(X)=Var(X)$ so the variance in the data can be better accounted for. So verify, we fit a model with negative binomial.
```{r echo=TRUE, warning=FALSE}
library(MASS)
model <- glm.nb(ER_visits ~ Drug , data=hd_data)
summary(model)
```
The residual deviance is 830.94 on 786 degrees of freedom, so it is a better model.
Q2(e)
```{r echo=FALSE, warning=FALSE}
results <- data.frame(Model = character(), P_Value = numeric(), Scaled_Deviance = numeric(), AIC = numeric())

  model <- glm.nb(ER_visits ~ Drug, data=hd_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "Drug", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm.nb(ER_visits ~ Interventions, data=hd_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "Interventions", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm.nb(ER_visits ~ logtotal_cost, data=hd_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "logtotal_cost", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm.nb(ER_visits ~ Age, data=hd_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "Age", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm.nb(ER_visits ~ Complications, data=hd_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "Complications", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm.nb(ER_visits ~ logcomorbidities, data=hd_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "logcomorbidities", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm.nb(ER_visits ~ Duration, data=hd_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "Duration", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
  model <- glm.nb(ER_visits ~ factor(Gender), data=hd_data)
  p_value <- summary(model)$coefficients[2, 4]
  scaled_deviance <- model$deviance / summary(model)$dispersion
  aic_val <- AIC(model)
  results <- rbind(results, data.frame(Model = "Gender", P_Value = p_value, 
                                       Scaled_Deviance = scaled_deviance, AIC = aic_val))
print(results)
```
Q2(f)
```{r echo=FALSE, warning=FALSE}
results <- data.frame(Model = character(), AIC = numeric())
model1<- glm.nb(ER_visits ~ Drug + Interventions + logtotal_cost + Age + Complications + Duration + factor(Gender), data = hd_data)
aic_val <- AIC(model1)
results <- rbind(results, data.frame(Model = "Drug + Interventions + logtotal_cost + Age + Complications + Duration + factor(Gender)", AIC = aic_val))
model2<- glm.nb(ER_visits ~ Drug + Interventions + logtotal_cost + Age + Complications + factor(Gender), data = hd_data)
aic_val <- AIC(model2)
results <- rbind(results, data.frame(Model = "Drug + Interventions + logtotal_cost + Age + Complications + factor(Gender)", AIC = aic_val))
model3<- glm.nb(ER_visits ~ Drug + Interventions + logtotal_cost + Age + factor(Gender), data = hd_data)
aic_val <- AIC(model3)
results <- rbind(results, data.frame(Model = "Drug + Interventions + logtotal_cost + Age + factor(Gender)", AIC = aic_val))
model4<- glm.nb(ER_visits ~ Drug + Interventions + logtotal_cost + factor(Gender), data = hd_data)
aic_val <- AIC(model4)
results <- rbind(results, data.frame(Model = "Drug + Interventions + logtotal_cost + factor(Gender)", AIC = aic_val))
model5<- glm.nb(ER_visits ~ Drug + logtotal_cost + Age + factor(Gender), data = hd_data)
aic_val <- AIC(model5)
results <- rbind(results, data.frame(Model = "Drug  + logtotal_cost + Age + factor(Gender)", AIC = aic_val))
model6<- glm.nb(ER_visits ~ Drug + logtotal_cost + factor(Gender), data = hd_data)
aic_val <- AIC(model6)
results <- rbind(results, data.frame(Model = "Drug  + logtotal_cost + factor(Gender)", AIC = aic_val))
print(results)
```

We use backward selection process starting from full model. Removing Duration reduces AIC, removing Complications reduces AIC. Removing anything else from the model increase AIC. So we choose model 3 as the final model.
Q2(g)
```{r echo=TRUE, warning=FALSE}
summary(model3)
```
ER_visits ~ NB(r,p)

The model equation is:
$$ log(ER_visits) = \beta_0 + \beta_1*Drug + \beta_2*Interventions + \beta_3*logtotal_cost $$
$$ + \beta_4*Age + \beta_5*Gender1 $$
here,
$$ \beta_0 = 0.1468, \beta_1 = 0.2144, \beta_2 = 0.0147, \beta_3 = 0.0632, \beta_4 = 0.007, \beta_5 = 0.1821 $$
Where,

Gender1 is a dummy variable. when gender is male, Gender1 is 1, 0 otherwise.

ER_visits is the response variable and remaining variables are continuous variables of the same name.

So the final model is,
$$ log(ER_visits) = 0.1468+0.2144*Drug+0.0147*Interventions+0.0632*logtotal_cost+0.007*Age+0.1821*Gender1 $$
Q2(h)
a 10% increase in total cost has the effect of $1.1^(0.0632)=1.0060$ increase in ER_visits or a 0.06% increase

The multiplicative effect on expected ER_visits of being male compared to not being male 
is $e^(0.1821)=1.1997$ times increase or 20% increase

The multiplicative effect on expected ER_visits of an increase in 1 prescribed Drug is $e^(0.2144)=1.2391$ times increase or 24% increase

The multiplicative effect on expected ER_visits of an increase in 1 Age is $e^(0.007)=1.0070$ times increase or .07% increase

The multiplicative effect on expected ER_visits of an increase in 1 in Interventions is $e^(0.0147)=1.0148$ times increase or 0.148% increase

Q2(i)
Diagnostic checks:
```{r echo=FALSE, warning=FALSE}
par(mfrow=c(1,1))
summary(model3)
```
scaled deviance 821.64 is fair against 782 degrees of freedom.
```{r echo=FALSE, warning=FALSE, fig.align='left'}

res_model3 <- simulateResiduals(fittedModel = model3)
q_model3 <- residuals(res_model3, quantileFunction = qnorm)
p1 <- ggplot(data = tibble(q = q_model3), aes(x = q_model3)) +
geom_histogram(aes(y = after_stat(density)), bins = 6) +
geom_density()
p1
```
quantile residuals look normally distributed

```{r echo=FALSE, warning=FALSE, fig.align='left'}
plot(res_model3)
```
The QQ plot does not show any significant results which is good for model fit.Residual vs predicted plot also does not detect any significant problem so the model is a good fit.
```{r echo=FALSE, warning=FALSE, fig.align='left'}
crPlots(model3)
```
All the variables look fairly linear. We can say systematic part of the model is specified correctly.
```{r echo=FALSE, warning=FALSE, fig.align='left'}
diag4 <- broom::augment(model3)
p1<-ggplot(diag4) +
geom_point(aes(y = .resid, x = seq_along(.resid))) +
xlab("Index") + ylab("Deviance Residuals")
```
Deviance residuals are a random scatter. no obvious pattern which is good. Leverage cut-off in this case is 0.01
```{r echo=FALSE, warning=FALSE, fig.align='left'}

par(mfrow=c(1,3))
p2<-ggplot(diag4, aes(y = .hat, x = seq_along(.hat))) +
geom_point() +
xlab("Index") +
ylab("Leverage") +
geom_hline(yintercept = 0.01, linetype = 2) +
ggrepel::geom_text_repel(aes(label = seq_along(.hat)))

p3<-ggplot(diag4, aes(y = .cooksd, x = seq_along(.hat))) +
geom_point() +
xlab("Index") +
ylab("Cook’s Distance") +
ggrepel::geom_text_repel(aes(label = seq_along(.hat)))
p1+p2+p3
```
We see quite a few high leverage points.
No cook's distance >1 so we conclude no unusual observation.

Q3(a) We can find the pdf of standard logistic distribution by differenciating the cdf function.
$$ f(u) = d/du(1 + e^(-u)) => f(u) = e^(-u)/(1+e^(-u))$$
Q3(b) To show that $\epsilon$ is symmetric about 0, we show that, $f(u) = f(-u)$
Q3(d) we can define a GLM of binomial distribution using the two equations. The link function is the logistic function.
